\chapter{Introduction}
\section{Background}


Refactoring is a routine activity in software development in which developers restructure code to improve readability, maintainability, and evolvability while preserving externally observable behaviour; traditional definitions describe it as a sequence of small, semantics-preserving transformations \cite{opdyke1992refactoring,fowler1999refactoring,fowler2018refactoring}, and subsequent surveys categorise these refactoring operations and the tools that support them \cite{mens2004survey}. In practice, refactoring is widely encouraged because poorly structured code becomes an obstacle when enhancing and maintaining code, and long-lived systems need continual adjustment (as argued in work building on Lehman’s laws and empirical studies of software maintenance). At the same time, refactoring can be costly: it consumes developer time, introduces risk, and is frequently intertwined with other work such as feature development and bug fixing rather than occurring as isolated “refactoring-only” tasks \cite{murphyhill2009howrefactor,kim2011apilevel}.

\section{Motivation}

Most research and tooling that evaluates refactoring focuses on \textit{outcomes} rather than the \textit{process}. A common approach is to measure refactoring success as a change in static quality indicators between a “before” and “after” state: reductions in code smells, improvements in cohesion/coupling, decreases in complexity, or related metrics to do with maintainability. Typical strands of work include metric-driven detection or prioritisation of candidate refactorings (e.g., using metric changes to separate refactoring-like edits from functional ones, or to choose between different refactoring options), as well as refactoring recommendation systems that rank alternatives by their impact on design metrics. More recent work expands the signals used for guidance by incorporating traceability alongside code metrics when ranking refactoring solutions, and exploring how product/process metrics relate to developers’ motivations to refactor (e.g., studies building on mined refactorings and repository history, and proposals that use LLMs to determine developers' motivations from commit artefacts). Alongside this, there is a substantial line of work that has improved the ability to \textit{detect} refactorings from code changes, notably by using AST-based differencing and structured matching to infer refactoring operations between revisions (e.g., RefactoringMiner and related approaches), which has enabled large-scale empirical studies of how often refactoring happens, what kinds occur, and in what contexts \cite{alikhanifard2025rminer,falleri2024gumtree}.

However, these perspectives evaluate refactoring as a mapping from an initial program state to a final program state, leaving the \textit{trajectory} between those states largely outside the picture. This matters because developers rarely refactor in a single atomic operation: they work through intermediate states, sometimes take detours, redo work, and often accept temporary degradation (e.g., in readability, modular structure, or test/build stability) before arriving at a final outcome. Two developers can arrive at the same end state via very different paths: one might take lots of disruptive steps, bounce between designs, or create avoidable churn, while another might get there via a steadier, more stable sequence. Existing endpoint-only evaluation cannot distinguish these different cases. Meanwhile, research on process realism suggests that refactoring is often ``flossed'' into ongoing development instead of being a separate dedicated task, which increases the chance of mixed-intent steps and noisy traces \cite{murphyhill2009howrefactor}. This makes the quality of the process (how a developer navigates intermediate states) potentially important for both developer productivity and the health of the codebase as it evolves.

Recent work has started to look more closely at refactoring trajectories themselves. RefactorBench, for example, uses a state/action/observation view to study how language-model agents fail on multi-file refactoring tasks \cite{gautam2025refactorbench}, while Bouzenia et al. characterise recurring action patterns in software-engineering agent traces \cite{bouzenia2025trajectories}. A recent systematic literature review also surveys how LLM-based refactoring work is evaluated and highlights several open problems in this area \cite{alomar2025slr}. However, this still leaves a gap for evaluating an observed developer trace directly. The closest related systems either synthesise a path to a desired end state, without judging the path the developer actually took, or recommend refactorings towards an architectural target, without reference to an observed process. This project addresses that gap by scoring recorded refactoring traces with an explicit process-quality metric, generating higher-scoring counterfactual reference trajectories, and identifying measurable points where the observed trace diverges from those references. 

\section{Research Objectives}

This thesis asks whether, given the start and end state of a refactoring session, we can judge the quality of the path the developer took and point to places where a better path was available. The core idea is to model refactoring as a \textit{process} over states, rather than judging it only by comparing endpoints. To do this, we adopt a state/action/trace framing: snapshots of the codebase are treated as \textit{states}, and developer edits-including IDE refactorings and manual transformations-are treated as \textit{actions} that produce a trace from the start state to the end state.

A core component will be a \textbf{process-quality metric} that scores traces according to criteria that will be defined and justified using existing research on code quality metrics, smell detection validity, and process-level signals. The metric is not assumed to be purely structural: it may incorporate multiple terms capturing endpoint improvement, intermediate-state degradation, churn/disruption, and “safety” proxies such as preserving build and test behaviour, reflecting the known limitations of relying on any one signal alone.

To go beyond scoring a single observed refactoring trace we also construct \emph{counterfactual} alternatives: \textbf{reference trajectories} between the same start and end states that score better under the chosen process-quality metric. These references give the observed trace something concrete to be compared against. The analysis can then identify \emph{divergence points}, where the developer's path moves away from a higher-scoring alternative, and estimate how much each divergence contributes to the final quality gap. This is related to systems that generate or recommend refactoring sequences: ReSynth synthesises a path to a target end state from a developer's partial edits \cite{raychev2013resynth}, Refactoring Navigator recommends steps towards an architectural target \cite{lin2016navigator}, and search-based refactoring treats sequences as an optimisation problem under explicit quality functions \cite{harman2007pareto,mohan2019manyobjective}. The difference is that these systems are mainly concerned with producing a useful sequence. Here, the generated trajectory is used as a reference for evaluating the process the developer actually followed.

\begin{figure}[H]
\centering
\begin{tikzpicture}[
  xscale=1.05, yscale=0.10,
  every node/.style={font=\small},
]
  % Axes
  \draw[->, thick] (-0.3, 35) -- (7.3, 35) node[right] {Step};
  \draw[->, thick] (-0.3, 35) -- (-0.3, 78) node[above] {$J$};
  \node[left] at (-0.3, 40) {\footnotesize 40};
  \node[left] at (-0.3, 60) {\footnotesize 60};
  \node[left] at (-0.3, 75) {\footnotesize 75};

  % Shared prefix (steps 0..3) — both trajectories follow the same path
  \draw[thick] (0,40) -- (1,45) -- (2,42) -- (3,50);

  % User's continuation (solid) — steps 3..6
  \draw[thick] (3,50) -- (4,55) -- (5,52) -- (6,60);
  \foreach \p in {(0,40),(1,45),(2,42),(3,50),(4,55),(5,52),(6,60)}
    \filldraw[black] \p circle (0.06 and 0.6);
  \node[anchor=west] at (6.1, 60) {$s_T^{\,\mathrm{user}}$};

  % Alternative trajectory (dashed, coloured to match the orange used
  % for divergent / counterfactual material elsewhere in the thesis).
  \draw[thick, dashed, orange!70!black] (3,50) -- (4,63) -- (5,60) -- (6,68);
  \foreach \p in {(4,63),(5,60),(6,68)}
    \draw[thick, orange!70!black, fill=white] \p circle (0.06 and 0.6);
  \node[anchor=west, orange!70!black] at (6.1, 68) {$s_T^{\,\mathrm{alt}}$};

  % Divergence-point marker
  \draw[->, gray, thick] (3, 30) -- (3, 49);
  \node[gray, anchor=north] at (3, 30) {\footnotesize Divergence point};

  % Gain bracket between the two terminal scores
  \draw[decorate, decoration={brace, mirror, amplitude=4pt}, gray]
    (7.30, 60) -- (7.30, 68);
  \node[gray, anchor=west] at (7.40, 64)
    {\footnotesize $\Delta J$};

  % Legend
  \draw[thick] (0.2, 76) -- (1.0, 76);
  \node[anchor=west] at (1.0, 76) {\footnotesize User trajectory};
  \draw[thick, dashed, orange!70!black] (3.6, 76) -- (4.4, 76);
  \node[anchor=west] at (4.4, 76) {\footnotesize Alternative trajectory};
\end{tikzpicture}
\caption{Illustration of two refactoring trajectories scored by the process-quality metric $J$. The solid line shows the developer's observed trajectory, while the dashed line shows a higher-scoring reference trajectory that diverges at an intermediate state. The gap $\Delta J$ is the estimated contribution of that divergence to the final score difference.}
\label{fig:divergence}
\end{figure}

\section{Contributions}

The main contributions of the project are:

\begin{itemize}
  \item A process-quality metric for refactoring trajectories. The metric combines terms from prior work on code quality and developer process signals -- including both endpoint cleanliness gain and an intermediate-cleanliness lag penalty so that two trajectories which finish at the same code can still score differently if one front-loaded the cleanup -- and we test how much the results change under sensitivity and ablation experiments.
  \item A divergence-point detector that compares an observed trace with a higher-scoring reference trajectory. The detector groups divergences into four kinds: \emph{Ordering}, \emph{Manual-Refactor}, \emph{Rework}, and \emph{Hygiene}.
  \item A 45-session dataset of recorded refactoring sessions on a purpose-built Java codebase. We provide multi-label ground-truth annotations and the experimental scripts used for the sensitivity, ablation, and divergence analyses.
  \item An empirical evaluation of the detector, reporting precision and recall for each divergence kind. The evaluation also shows which parts of the score formula are useful in practice, and which kinds of process improvement it does not capture well.
\end{itemize}

\section{Approach and Scope}

The approach is intended for realistic refactoring sessions, where IDE-supported refactorings and manual edits are often mixed together. In this project we focus on Java, since it gives us mature parsing tools, established quality metrics, and refactoring mining support such as RefactoringMiner. We represent a refactoring session as a sequence of code snapshots. Each step may correspond to a burst of manual edits or the use of an automated IDE refactoring tool, and the same model could later be applied to traces at different levels of granularity where that data is available.

The work does not attempt to enumerate every possible refactoring sequence or prove that a generated trajectory is globally optimal. Instead, it uses the analysis component to find a plausible higher-scoring reference path between the same start and end states. This keeps the task focused on explaining the developer's observed process, rather than replacing the developer with an automated refactoring planner.

The work is scoped to human developer sessions. The agent comparison in Chapter~\ref{ch:results} is a feasibility extension on the same tooling rather than a co-equal evaluation target: the detectors are designed against human edit patterns (IDE refactor invocations, edit-burst structure, commit-cadence windows) and do not map cleanly onto agent traces. We treat that mismatch itself as one of the things Chapter~\ref{ch:results} surfaces.

\section{Thesis Roadmap}
The remainder of the thesis is organised as follows. Chapter~2 reviews related work on refactoring evaluation, refactoring mining, and refactoring recommendation. Chapter~3 defines the process-quality metric and divergence-point detector. Chapter~4 describes the implementation of the tool. Chapter~5 presents the evaluation and discusses threats to validity. Chapter~6 concludes the thesis and outlines possible future work.
