\chapter{Discussion}
\label{ch:conclusion}

This chapter brings together the main results of the project. \S\ref{sec:conclusion-promise} highlights the parts of the work that landed well. \S\ref{sec:conclusion-contributions} summarises what was built. \S\ref{sec:conclusion-limitations} discusses the main limitations. \S\ref{sec:conclusion-findings} states the headline findings, and \S\ref{sec:conclusion-future-work} covers directions for future work.

\section{Signs of promise}
\label{sec:conclusion-promise}

The results give early evidence that process-level refactoring evaluation is a viable direction. We did not fit the score formula to a labelled outcome dataset, yet its highest-ranked recommendations are generally stable under weight perturbation: on the user-study sessions, the top-ranked divergence point is preserved in $96.5\%$ of single-weight perturbation cases, and under joint perturbations the top recommendation remains stable in $84.6\%$ of samples (\S\ref{sec:results-sensitivity}). The main feedback the tool surfaces is therefore not simply an artefact of one fragile weight setting, although lower-ranked points are more sensitive to simultaneous changes across several weights. The ablation results also show that the process terms are doing real work: length, manual-when-IDE, broken-build time, skipped tests, and commit gaps each affect the recovered magnitude or ranking in different settings (\S\ref{sec:results-ablation}), which supports the central claim that refactoring-process quality cannot be reduced to endpoint cleanliness alone.

The divergence detector also looks promising as an explanatory mechanism, not just a scoring layer. On the 45-session labelled injection set, precision is $1.00$ for all four divergence kinds, with perfect recall for Rework and Hygiene, strong recall for Manual-Refactor, and weaker recall for Ordering because the reorder synthesiser rejects windows it cannot safely reproduce (\S\ref{sec:results-divergence}). The surfaced alternatives are also often useful in their own right: $26$ of the detected injection cases produce a positive-magnitude alternative, and every positive-magnitude catch is ranked first within its session. The user study adds a smaller but encouraging end-to-end signal. Fitting a line through the per-participant gain-stripped scores gives a $+4.47$ per-session slope for the with-feedback group, against $-0.40$ for the no-feedback group; across all $30$ sessions the with-feedback group also produces divergence points at $2.7$ per session, against $6.0$ per session for the no-feedback group, a $2.2\times$ difference (\S\ref{sec:results-userstudy}). At $n = 3$ vs $n = 2$ this is directional rather than conclusive, but it is still a useful sign that the feedback may have changed how participants approached later refactoring sessions.

\section{Summary of contributions}
\label{sec:conclusion-contributions}

The project makes two main contributions. The first is the process-quality score $J$. It combines the endpoint cleanliness gain with five process penalties (broken builds, skipped tests, manual edits where an IDE refactoring was available, long gaps between commits, and an intermediate-cleanliness lag penalty) and a step-savings bonus for alternatives that reach the same end state in fewer steps. The second contribution is the divergence-point detector. It finds points in a refactoring session where a better alternative exists, assigns them to one of four kinds (\emph{Manual-Refactor}, \emph{Rework}, \emph{Hygiene}, \emph{Ordering}), and produces a concrete alternative trajectory for each. The alternatives come from four synthesisers: one reorders steps, one replays manual edits as IDE refactorings, one removes self-cancelling rework, and one models different test and commit choices.

The work is supported by a 45-session hand-recorded dataset on a purpose-built Java fixture with multi-label ground-truth annotations, a 30-session randomised-group user study on a separate fixture, and an 8-agent, 48-session extension run on the same user-study fixture. It also includes the IntelliJ plugin used to capture event traces, the JCEF dashboard used to surface divergence points, and a reproducible notebook from which every number in Chapter~\ref{ch:results} can be regenerated.

\section{Limitations}
\label{sec:conclusion-limitations}

This section sets out the main limitations to keep in mind when reading the results in Chapter~\ref{ch:results} and the headline findings in \S\ref{sec:conclusion-findings}. Some mitigations are already described alongside the relevant design choices: the three-way inter-rater check, deterministic ranking tie-breaks, the multi-knob sensitivity analysis, and the reproducible notebook used for the reported numbers. They are only repeated here where they directly affect the limitation being discussed.

\subsection{Construct and weight calibration}
\label{sec:threats-construct}

\paragraph{There is no external ground truth for ``refactoring quality''.}
The process-quality score $J$ measures signals that suggest a weaker refactoring process: broken builds, skipped tests, manual edits where an IDE refactoring would have worked, and long gaps between commits. It should not be read as a predictor of downstream outcomes such as later bug rates or time-to-merge. No dataset links session-level process signals to maintenance outcomes at this granularity, so the weights follow the severity orderings described in \S\ref{sec:methodology-weights} rather than being fitted to outcome data.

\paragraph{The weights are plausible, but not uniquely calibrated.}
The production weights follow the severity ordering described in the methodology, but the exact values are still a design choice, and other values within the same ordering could also be justified. The sensitivity analysis in \S\ref{sec:results-sensitivity} gives some reassurance: on the rankable user-study subset ($25$ of $30$ sessions), the top-ranked divergence point stays the same in $96.5\%$ of single-knob perturbations and $84.6\%$ of multi-knob Monte Carlo runs. The formula is therefore not fragile near the chosen weights, but this is still only a local robustness result, not a claim about arbitrary reweighting.

\paragraph{Agreement was checked on kinds, not ranking.}
The Cohen's $\kappa$ result in \S\ref{sec:results-kappa} shows that three independent labellers usually agree on which divergence kinds appear in each session. It says nothing about whether the same labellers would rank those divergence points in the same order as the score, even though the dashboard presents the highest-ranked point first. Checking that would require a separate ranking study, for example comparing human rankings against the production ranking using Kendall $\tau$-b. I leave this as future work in \S\ref{sec:conclusion-future-work}.

\subsection{Study design and statistical power}
\label{sec:threats-study-design}

\paragraph{The user study is randomised, but too small for hypothesis testing.}
Five participants completed six sessions each, for $30$ sessions in total. Group assignment was uniformly random before the study began, but the groups themselves are very small ($n = 3$ vs $n = 2$). Within-group variation is about the same size as the between-group difference, so the results chapter reports only the direction of the effect, without a $p$-value or confidence interval. The no-feedback group used the same playbook, and its downward trend makes it less plausible that the tasks simply became easier over time. Two alternative explanations still remain. Participants in the with-feedback group may have learned what the dashboard rewards, or they may have become more comfortable with the session format rather than better at refactoring itself.

\paragraph{The author is one of three raters.}
One of the three labellers was me, so my own judgement may have affected the agreement results. To make this easier to inspect, \S\ref{sec:results-kappa} reports pairwise $\kappa$ rather than only a single aggregate score. The other two labellers were later placed in the with-feedback group of the user study, but they completed all labels before taking part in any study sessions, so their labels could not have been shaped by their later behaviour in the experiment. The remaining concern goes the other way: two of the three with-feedback participants had already seen the divergence-point taxonomy during labelling. For that reason, \S\ref{sec:results-userstudy} reports per-participant trajectories rather than burying this inside the group mean.

\subsection{Generalisation and implementation scope}
\label{sec:threats-external}

\paragraph{The injection sessions are controlled exercises.}
The $45$ injection sessions in the primary test set follow a scripted playbook, with one specific bad behaviour added to each. This makes them useful for checking whether the detector finds known divergence kinds, but they are not typical refactoring sessions. The precision and recall figures in \S\ref{sec:results-divergence} should therefore be read as controlled-test results, not as direct estimates for ordinary development work.

\paragraph{\texttt{SuboptimalOrdering} is based on the playbook author's judgement.}
The five ordering-injection sessions are labelled \texttt{SuboptimalOrdering} because the playbook author chose an order that seemed worse than the available alternatives. The score formula only agrees with that judgement in $1$ of the $5$ sessions (\S\ref{sec:results-divergence}, session~$022$); the other $4$ have zero or negative magnitude under the production formula. The \emph{Ordering} precision and recall numbers should therefore be read with this mismatch in mind.

\paragraph{The agent extension is a feasibility experiment, not an agent evaluation.}
The eight-agent, $48$-session extension tests how detectors built for human IDE sessions behave on a different kind of actor. It is not an evaluation of the tool as an agent-monitoring framework. Three of the four detector kinds fit agent traces poorly (\S\ref{sec:results-userstudy-agent-limits}). Agents have no IDE refactoring surface, so \emph{Manual-Refactor} fires on every structural refactoring without offering a realistic alternative. The six-checkpoint commit-gap threshold is rarely reached within six sessions, and the $2$-second edit-burst debouncer collapses each agent response into one burst. The agent result should therefore not be read as evidence that feedback had no effect on agent behaviour. It mainly shows that the current tool is not well matched to agent edit patterns.

\paragraph{The reorder enumerator has a hard size budget.}
The \emph{Ordering} detector skips reorder windows with more than $7$ refactoring steps, because enumerating all valid orderings becomes expensive past that point (\S\ref{sec:methodology-reorder}). The recorded sessions rarely reach this limit, but a session with a longer uninterrupted chain of refactorings would lose its largest windows from the \emph{Ordering} analysis.

% Backward-compatible anchors for cross-references from earlier chapters.
\label{sec:threats-internal}
\label{sec:threats-statistical}
\label{sec:threats-per-experiment}
\label{sec:threats-implementation}
\label{sec:threats-out-of-scope}

\section{Headline findings}
\label{sec:conclusion-findings}

With those limitations in mind, four main findings come out of the project.

First, the kind classifier performs well on the controlled-injection sessions. Against the inter-rater-agreed \texttt{expected\_kinds} labels, per-kind precision is $1.00$ for all four kinds. The agreement check is also strong: Cohen's $\kappa = 1.00$ for \emph{Ordering} and \emph{Manual-Refactor} on all three rater pairs, and $\kappa = 0.72$--$1.00$ for \emph{Rework} and \emph{Hygiene} (\S\ref{sec:results-kappa}, \S\ref{sec:results-divergence}). Every detected injection with a positive-magnitude alternative trajectory ranked at top-$1$ within its session.

Furthermore, the score formula is locally stable around its production weights. On the rankable user-study sessions ($25$ of $30$), changing one weight at a time preserves the top-ranked divergence point in $96.5\%$ of cases. Perturbing all weights at once lowers this to $84.6\%$ (\S\ref{sec:results-sensitivity}). The formula is therefore not fragile near the chosen weights, although the result is still local: larger or differently structured changes to the weights can change the rankings.

Also, the user-study group comparison points in a clear direction. Fitting a line through the six per-participant gain-stripped scores gives a $+4.47$ per-session slope for the with-feedback group, against $-0.40$ per session for the no-feedback group. Across all $30$ sessions, the with-feedback group also produces divergence points at a rate of $2.7$ per session, against $6.0$ per session for the no-feedback group, a $2.2\times$ difference. With $n = 3$ vs $n = 2$, the study is randomised but small, so no $p$-value is claimed; the result is reported only as a direction of effect. The two groups start at comparable totals in S1 and diverge from S3 onwards, which makes it harder to argue that the with-feedback improvement simply reflects the playbook getting easier over time (\S\ref{sec:results-userstudy}).

Finally, the agent extension is best read as a feasibility experiment, not as a full evaluation of the tool on agents. It covers eight agents across model $\times$ harness combinations, with six feedback agents, two no-feedback baselines, and $48$ sessions in total. The detectors were built around human IDE sessions, and they do not transfer cleanly to agent traces: of the $42$ divergence points found across the agent sessions, $41$ are \emph{Manual-Refactor}, because agents have no IDE refactoring surface to switch to and the other three detector kinds rarely fire (\S\ref{sec:results-userstudy-agent-limits}). The feedback condition also shows no clear effect on agent process scores. Both groups have positive slopes, but the no-feedback baseline slope is larger ($+1.3$ versus $+0.6$), so the visible difference runs in the opposite direction to the human result. The transcripts add context that the score does not capture: five of the six feedback agents explicitly write down plans for what they will change next, but the gain-stripped $\Delta J$ still does not separate from the no-feedback group. This fits the detector-scope limitation, since most agent adjustments concern build-broken time and commit cadence, whereas the measured agent sessions are dominated by \emph{Manual-Refactor}.

\section{Future work}
\label{sec:conclusion-future-work}

The most useful next steps are the ones that would test the main findings against stronger evidence:

\begin{enumerate}
    \item \textbf{Larger randomised human study.} The strongest evidence for a feedback effect is the with-feedback vs no-feedback group comparison in \S\ref{sec:results-userstudy}, but it rests on only five participants. A larger study would show whether the same direction holds with more people, and would make it possible to estimate between-participant variance properly.

    \item \textbf{Adapting the tooling to agent traces.} The current detectors are built around human IDE behaviour, so they do not fit agent traces cleanly. A version aimed at agents would need different assumptions: grouping edits by tool call rather than by short edit bursts, tracking commits per agent response rather than through the six-checkpoint window, and checking which refactoring tools the agent can actually use. That would let the same scoring framework be tested on agents without the current detector mismatch.

    \item \textbf{IDE refactoring tools for agents.} The clearest agent-side limitation is that feedback can recommend an IDE refactoring, but the agent has no way to invoke it. Exposing IntelliJ refactorings as agent tool calls would make that feedback actionable, and would allow the \emph{Manual-Refactor} count to change across sessions.
\end{enumerate}

\section{Closing}
\label{sec:conclusion-closing}

This thesis shows that the path a developer takes through a refactoring session can be tracked, scored, and compared against concrete alternative paths that score better on a specific process metric, with the gap between observed and alternative localised at identifiable points. The score formula, detector, and per-kind synthesisers developed here remain limited by their calibration, by the size of the dataset, and by what the four detector kinds can see. Even so, controlled divergences are recovered with per-kind precision $1.00$, the rankings stay stable in $96.5\%$ of single-knob perturbations and $84.6\%$ of multi-knob perturbations near the chosen weights, and the randomised user-study arm comparison points in a clear direction. The agent extension shows where the same framework needs adapting for traces that do not follow human IDE patterns, which is itself a useful outcome: it isolates the detector assumptions that are human-specific. What remains is to test the framing on larger human studies, on broader refactoring support, and with detectors built specifically for agent traces.
