\chapter{Threats to validity}
\label{ch:threats}

This chapter collects the limitations mentioned throughout Chapter~\ref{ch:results} and discusses design choices that affect the validity of our claims. The discussion follows the standard four-axis framework from empirical software engineering~\cite{wohlin2012experimentation}: construct validity, internal validity, external validity, and statistical conclusion validity. A per-experiment section then identifies threats specific to each of the four empirical studies (the kind-classifier, the user study, the agent comparison, and the rater study), followed by implementation-side concerns and an out-of-scope list. The mitigations the design already carries (three-way inter-rater reliability on the injection sessions, deterministic ranking tie-breaks, multi-knob robustness alongside the single-knob sweep, honest score-floor and validator-check disclosures, reproducible notebook sources for every reported number, and an explicit per-session-not-per-step true-positive policy) are stated where the corresponding design choice or result is presented, rather than restated here.

\section{Construct validity}
\label{sec:threats-construct}

\paragraph{There is no external ground truth for ``true refactoring quality''.}
The process-quality score $J$ measures process behaviours the literature identifies as bad practice: leaving the build broken, going long stretches without committing, refactoring manually where an IDE refactor would have worked, and so on. It does not predict bug rates, time-to-merge, or any other downstream-of-refactoring outcome, and Chapter~\ref{ch:methodology} does not claim it does. Calibrating a formula like this against a held-out outcome dataset would require a dataset linking session-level process signals to those outcomes, and we are not aware of one that exists. The weights are grounded in the literature-supported severity orderings of \S\ref{sec:methodology-weights} rather than fitted to data; the chapter addresses robustness and inter-rater agreement empirically, but not predictive validity.

\paragraph{The score is not a maintainability index.}
Snapshot composites like the Maintainability Index~\cite{coleman1994maintainability} were regression-fitted against expert ratings of completed code. Our score is trajectory-level and process-side; it asks ``did the refactoring follow good practice on the way there'', not ``is the code good now''. The two are different constructs and we do not conflate them.

\paragraph{The four-kind taxonomy is a methodology choice, not a discovered structure.}
The divergence-point kinds (\emph{IDE-Replay}, \emph{Rework}, \emph{Hygiene}, \emph{Ordering}) were chosen for the rater-design rationale set out in \S\ref{sec:methodology-rater}: they are operationally distinguishable behaviours the literature flags as bad practice, and they admit a synthesiser that can produce a concrete alternative for each. A different taxonomy is defensible. Collapsing \emph{Hygiene} into separate top-level \emph{Tests-Skipped} and \emph{Commit-Gap} kinds, or merging \emph{Rework} into \emph{Ordering} on the grounds that both are about wasted steps, would change the precision/recall numbers without changing the chapter's headline claims.

\paragraph{\texttt{SuboptimalOrdering} is author-claimed, not measured.}
The five ordering-injection sessions in the playbook carry the \texttt{SuboptimalOrdering} pattern label because the playbook author selected an ordering they believed was worse than its alternatives. The score formula cannot independently verify all five: only $1$ of the $5$ ``suboptimal'' orderings produces a positive-magnitude reorder alternative under the production formula (\S\ref{sec:results-divergence}, session $022$), while the other $4$ remain at magnitude zero because their permutations commute and the synthesiser cannot reach an AST-distinct end state. The label reflects what we designed the session to contain rather than objective ground truth, so interpret the precision/recall numbers with that in mind.

\paragraph{The AST-equivalence audit uses an informal definition.}
A candidate permutation is accepted by the reorder synthesiser only if its terminal AST is equivalent to the user's. Equivalence is defined as ``modulo syntactic differences that do not change behaviour'' (\S\ref{sec:methodology-reorder}) and implemented using an AST hash. The exact set of differences the hash treats as behaviour-preserving (e.g.\ comment placement, annotation order, whitespace) is implicit in the hash implementation rather than spelled out here. A hash that is too loose would admit reorders that change behaviour and a hash that is too tight would reject reorders that are genuinely equivalent. It catches the obvious failures, but the exact boundary is a design choice we do not formally specify.

\section{Internal validity}
\label{sec:threats-internal}

\paragraph{The single-knob sensitivity sweep is combined with a multi-knob design rather than replaced.}
The headline top-$1$ stability number in \S\ref{sec:results-sensitivity} is computed under single-knob perturbation: one weight changes per row, the rest stay at production. Real misspecification of weights rarely happens one at a time, so a multi-knob Monte Carlo experiment runs alongside (every weight perturbed simultaneously from $\exp(\sigma \cdot \mathcal{N}(0, 1))$ with $\sigma = \ln 2$). Top-$1$ stability falls from $96.5\%$ under single-knob to $84.6\%$ under multi-knob on the user-study rankable subset ($25$ of $30$ sessions). The formula is robust to plausibly-sized perturbations of any single weight, and robust to roughly five-in-six plausibly-sized simultaneous perturbations of all weights, but is not robust to arbitrarily-large simultaneous moves. The rank statistics are computed on the rankable subset of each set ($15$ injection, $25$ user-study, $20$ agent) because $\tau$-b and top-$1$ are mechanically $1.0$ on sessions with $\leq 1$ detected divergence point; the smaller rankable subset on the injection set is a known statistical-power limit on cross-set rank comparison, though the §\ref{sec:results-ablation} magnitude-recovery statistics use the full $45$ injection sessions where the small-ranking issue does not apply.

\paragraph{Kind-presence agreement does not imply rank-order agreement.}
The Cohen's $\kappa$ reported in \S\ref{sec:results-kappa} establishes that three independent labellers agree on which divergence kinds fire in each session. It does not establish that they agree on the order the score puts those divergence points in, which is what the dashboard surfaces first to the user. Closing this gap would require a separate rater-rank study (engage raters to rank the divergence points within each session, compute Kendall $\tau$-b against the score's production ranking). This was out of scope for our project. Our strongest claim from $\kappa$ alone is that the score's kind labels match what the rater identifies, but the ranking that orders those labels is not externally validated.

\paragraph{Cleanliness sub-weights are uniform $1.0$ by Laplace's principle of insufficient reason.}
The six sub-signals that feed the cleanliness sub-score (\S\ref{sec:methodology-cleanliness}) are weighted equally because no published calibration data exists for weighting these signals across refactoring trajectories. The sensitivity sweep shows the cleanliness sub-weights have essentially no effect on the top of the divergence-point ranking on the user-study set (Table~\ref{tab:results-sensitivity-knobs}: five of six sub-knobs produce zero top-$1$ disruptions, and \texttt{cohesion} produces a single disruption attributable to the lag-denominator coupling). On the agent sessions the cleanliness sub-weights' Kendall $\tau$ drops by 3--10\,\% under $\times 0.5$/$\times 1.5$ perturbation -- the same coupling running through a smaller session set where one shifted rank dominates. The uniform choice is defensible in practice on the current sessions but it is not a positive calibration result.

\paragraph{The lag term couples the cleanliness sub-score to the process score.}
The lag term $W_{\textsc{lag}}\,\ell(\tau)$ uses the trajectory's final cleanliness scalar $C(s_T)$ as the reference against which intermediate-state deficits are measured (\S\ref{sec:methodology-cleanliness}). This means a perturbation of any cleanliness sub-weight propagates into the process score through both the endpoint-gain term and the lag denominator, instead of only through the endpoint. The agent-session sensitivity result above is the empirical signature of this coupling. The injection and user-study sessions absorb the same leakage without measurable rank disruption because their per-fixture rankings are longer and more diverse, but the underlying coupling is the same. A cleanliness composite that mis-weights one sub-signal could in principle shift the lag-driven part of the ranking by a small amount we cannot disentangle from the gain-driven part. We treat this as an acceptable design cost: the lag term is meaningful only in the cleanliness-scalar units used elsewhere, and detaching it would introduce a second scalar with no independent calibration anchor.

\paragraph{Two batching primitives are heuristic and not sensitivity-tested.}
The score formula and the divergence-point detector share two batching constants: \texttt{min\_commit\_gap}~$=6$ (the green-refactor checkpoint count at which a \emph{Commit-Gap} divergence point fires and the score formula applies the commit-gap penalty), and the $60$-second composite-window boundary that bundles refactoring steps into batches for \emph{Tests-Skipped} and \emph{Ordering} detection. Both come from Murphy-Hill et al.'s 2009 study of how developers actually refactor~\cite{murphyhill2009howrefactor}, and neither is sensitivity-tested directly. The sensitivity sweep perturbs the weights on these terms but not the thresholds themselves. A different typing speed (slower or faster than the participants in this study) would group steps differently and could produce a different mix of fired kinds.

\paragraph{The production weight vector is hand-picked within a literature-grounded ordering.}
The exact production weights (\S\ref{sec:methodology-weights}, Table~\ref{tab:methodology-weights}) sit within the severity ordering the literature supports ($W_b > W_{st} > W_{cg}$ etc.), but no calibration step pins them to those specific numbers rather than any other vector inside the same ordering. The robustness experiments establish that the formula sits in a locally stable region around this vector; they do not establish that no other defensible vector exists in the same region.

\paragraph{Which terms carry the ranking varies by session type.}
The ablation magnitude analysis on the full $45$ injection sessions (Table~\ref{tab:results-ablation-loo}) identifies \texttt{length}, \texttt{manualIde}, and \texttt{broken} as the strongest magnitude-side contributors (removing each costs sum/sum between $0.18$ and $0.28$). The cross-set rank-stability analysis on the rankable subsets (Table~\ref{tab:results-ablation-user-loo}) identifies \texttt{length} as the dominant rank-carrier on the user-study set ($\tau = 0.527$ on removal), with \texttt{manualIde} and \texttt{broken} next. The lag term has the smallest user-study rank impact ($\tau = 0.784$ on removal) but a measurable injection-set impact ($\tau = 0.409$), so its rank-shaping role is narrower on the naturalistic user sessions than on the controlled-injection set. The formula's effective parameter dimension therefore depends on what kind of session is being scored. This connects to the gain-stripped finding in \S\ref{sec:results-userstudy}: the cleanliness-gain term dominates the absolute score but does not reshape the divergence-point ranking on the user-study set, so the formula effectively has two modes (an outcome-side mode that drives the absolute level and a process-side mode that drives the ranking). Whether a weight matters depends on which mode is being read - the absolute score or the ranking.

\paragraph{The reorder dependency analysis is deliberately coarse, and anchors are content-addressed.}
The dependency analyser that decides which refactoring steps can be safely reordered uses two design choices that affect the \emph{Ordering} recall reported in \S\ref{sec:results-divergence}. First, \texttt{Extract*} and \texttt{InlineMethod} specs are modelled optimistically: two extracts on the same host method, or two inlines of distinct methods on the same class, are not auto-conflicted, on the grounds that the downstream AST-equivalence check will reject any reorderings that conflict in practice. Second, the four specs that rewrite call sites of a named method (\texttt{RenameMethod}, \texttt{ChangeMethodSignature}, \texttt{ParameterizeVariable}, \texttt{ParameterizeAttribute}) are modelled coarsely: each writes the entire declaring class, which chains every later operation on the same class after it by construction, even when the operations would have been genuinely independent. Other rename specs (\texttt{RenameClass}, \texttt{RenameField}, \texttt{RenameLocalVariable}) only consume and produce the specific named entity rather than writing the whole class. On top of this, the entities that anchor each refactoring (a specific region of code, a specific local declaration) are identified by AST subtree hashes, so an identifier rename inside an anchor's region between two reorderable steps invalidates the second step's anchor. The optimistic model over-enumerates but the AST-equivalence check filters out the resulting false positives; the coarse model under-enumerates and removes some legitimate independence by construction. The $0.36$ \emph{Ordering} recall number reflects these two design choices. The observation that reordering rarely improves the score under the production formula is robust to them, because that observation follows from commuting refactorings producing identical end states under the formula (\S\ref{sec:methodology-reorder}), which does not depend on how the dependency analyser decides what commutes.

\paragraph{Cleanliness sub-signals are clamped to the user's range on alternative trajectories.}
Alt-trajectory cleanliness sub-signals are min-max normalised against the main (user) trajectory's observed range, then clamped to $[0, 1]$. An alternative whose intermediate checkpoints sit outside the user's range is masked at the boundary. We mention the clamp in \S\ref{sec:methodology-cleanliness} but do not flag it as a threat there; it is named here for completeness. The clamp does not affect the divergence-point magnitudes we report because the validator requires every accepted alternative to reach the same terminal AST as the user's trajectory, so alt and user share the same terminal-state cleanliness by construction; the trajectory-final $J$ computed from $\Delta C = C(s_T) - C(s_0)$ is identical on both. The clamp only changes how alt intermediate checkpoints are displayed on the dashboard's cleanliness chart, which has no downstream effect on the scoring numbers.

\section{External validity}
\label{sec:threats-external}

\paragraph{The injection sessions are scoped exercises, not naturalistic sessions.}
The $45$ sessions in the primary test set were recorded by us against a scripted playbook that injects one specific bad behaviour per session. They are scoped exercises rather than naturalistic refactoring activity; precision and recall claims drawn from this dataset are conditional on that controlled-injection structure, not on the variety of session shapes a free-form developer would produce in practice.

\paragraph{The user study has five participants split into a randomised three-vs-two arm.}
Five participants performed six recorded sessions each, $30$ in total. Arm assignment was randomised before the study began: two of the five participants were drawn uniformly at random to form the no-feedback baseline arm, and the remaining three form the with-feedback arm. The chapter reports per-participant trajectories and arm-mean comparisons but does not report statistical tests, confidence intervals, or claims like ``the median participant improved on metric $X$''. The randomised design rules out the alternative reading that the with-feedback arc reflects the playbook getting easier across the arc - the baseline arm worked the same playbook in the same order and trended downward - but $n = 3$ vs $n = 2$ leaves the per-participant variance comparable to the between-arm difference. The strongest claim drawn from the arm comparison is therefore directional, not inferential: under random arm assignment, the with-feedback arm improved across the arc, the baseline arm did not, and the within-arm variance is smaller than the between-arm difference. The remaining confounds after that arm split - demand characteristics and task-style familiarity - are discussed under per-experiment threats in \S\ref{sec:threats-per-experiment}.

\paragraph{The agent extension is a feasibility experiment, not an agent evaluation.}
Eight agent stacks (six with-feedback, two no-feedback baselines) ran the same six-session playbook on the same fixture, $48$ sessions in total. The agent corpus exists to observe how the human-trained detectors behave on a structurally different actor, not to evaluate the tool as an agent-monitoring framework. Three structural detector biases against agent traces are documented in \S\ref{sec:results-userstudy-agent-limits} and listed again in the construct-validity item above: agents have no IDE refactor surface, so the \emph{IDE-Replay} kind fires on every structural refactoring without recourse to a fix; the six-checkpoint commit-gap threshold is rarely reached in a six-session arc; and the 2-second edit-burst debouncer collapses each agent turn into one burst. The arm-level $\Delta J$ comparison and the per-stack DP totals are therefore reported as descriptive feasibility evidence rather than as a verdict on the tool's intended use, which is human-developer refactoring. A multi-stack agent study under a detector set reworked for agent edit patterns is the natural follow-up; that work is named in \S\ref{sec:threats-out-of-scope} and the conclusion's future-work section.

\paragraph{All sessions are single-author IntelliJ sessions on a Gradle Java project.}
The detector's behaviour on team pull-request-driven workflows, on other IDEs, or on other languages is not studied. The plugin's instrumentation is IntelliJ-specific: it uses the IDE's PSI and VFS hooks to capture refactoring events, which have no direct equivalent in VSCode or Eclipse, so a port would require re-implementing the event-capture layer rather than just retargeting a build.

\paragraph{The detectors are designed for human edit patterns.}
The score and divergence-point detectors are built around how human developers interact with an IDE: discrete refactor-menu invocations, edit bursts that the plugin debounces on a 2-second idle, and commit cadences thresholded against the 60-second composite window of \cite{murphyhill2009howrefactor}. The agent sessions in Chapter~\ref{ch:results} are processed by the same pipeline, but the detectors do not map cleanly onto agent traces. Tool-applied file edits collapse onto a single burst because the model's reasoning pause between turns exceeds the 2-second debounce; the six-checkpoint commit-gap threshold is rarely reached in a six-session arc because each agent turn batches more file changes than a typical human edit; and the \emph{IDE-Replay} kind is structurally moot because the agent has no refactor-menu surface to choose between. The agent comparison is treated as a feasibility extension on the tooling rather than a target the detectors were tuned for.

\paragraph{Refactoring kinds the synthesiser cannot handle are excluded by construction.}
The \emph{IDE-Replay} synthesiser builds alternative trajectories by replaying refactoring specs through Eclipse JDT. A user trajectory built entirely from refactoring operations outside JDT's supported set (or outside RefactoringMiner's detection capability, which feeds the synthesiser) would produce no \emph{IDE-Replay} divergence points. The detector cannot flag what it cannot synthesise an alternative for. This is a structural limitation of the detection design, not a finding specific to the sessions we used.

\section{Statistical conclusion validity}
\label{sec:threats-statistical}

Cohen's $\kappa$ in \S\ref{sec:results-kappa} is reported as a point estimate with raw agreement counts. Confidence intervals are not reported because with only $45$ sessions per rater pair, they would be too wide to be useful and would overshadow the $\kappa$ value itself. Instead, the specific disagreement cases are listed (sessions 024 and 043 for \emph{Rework}; sessions 014, 015, 016, 039 for \emph{Hygiene}), which tells the reader more.

Kendall's $\tau$-b is brittle on small per-fixture rankings. A single pair swap on a three-point ranking moves $\tau$ by $0.333$. $\tau$ is reported on all three session sets, but the injection sessions' per-fixture rankings are short enough that the metric is dominated by the small-$N$ artefact; the sweep therefore reports top-$1$ hit rate alongside $\tau$ as a more stable surrogate, and the user-study sessions are treated as the canonical benchmark in \S\ref{sec:results-sensitivity}.

No multiple-comparison correction is applied across the weight $\times$ factor sweep ($12 \times 9 = 108$ perturbation cases per fixture). The headline result is a descriptive statistic - what fraction of cases preserve the top-$1$ recommendation - not a hypothesis test, so a correction would be appropriate only if individual cases were being interpreted as significance claims, which they are not. The multi-knob Monte Carlo uses a fixed seed and the $200$-sample size was chosen heuristically rather than from a power calculation; distribution shape (mean $\tau$, top-$N$ hit rate) is reported rather than a point estimate with confidence interval.

\section{Per-experiment threats}
\label{sec:threats-per-experiment}

\paragraph{The kind classifier (\S\ref{sec:results-divergence}).}
The $1.00$ precision figure across all four kinds is conditional on the sessions used. \texttt{expected\_kinds} is a hand-label computed independently of the playbook's injection pattern, and the inter-rater $\kappa$ study (\S\ref{sec:results-kappa}) establishes that three labellers agree on it for these sessions. The threat is not that the labels are author-biased; it is that the dataset consists of controlled-injection sessions whose divergence behaviours are bounded by what the playbook injected and what the four detector kinds can describe. A free-form session set could contain bad behaviours outside those four kinds, and even within the four kinds the labelling shows rater variance on this dataset ($\kappa = 0.72$ on \emph{Hygiene} between two of the three pairs, \S\ref{sec:results-kappa}). Precision on a free-form session set is therefore unstudied. Separately, IDE-replay precision was $0.80$ before the plugin-instrumentation fix described in \S\ref{sec:results-scoped-out}: the false positives in the pre-fix recordings were upstream plugin-capture issues, not detector-logic errors, but the current $1.00$ figure would not hold on the pre-fix recordings.

\paragraph{The user study (\S\ref{sec:results-userstudy}).}
The external-validity limitations above ($n = 3$ vs $n = 2$, randomised arm assignment) apply here too. The metrics in the gain-stripped score (commit cadence, skipped tests, manual-when-IDE rate) don't depend much on knowing the codebase, so the with-feedback arm's improvement can't be explained by just getting more familiar with the fixture - and the baseline arm worked the same fixture on the same playbook and trended downward, which removes ``the playbook got easier as the participant progressed'' as a competing explanation. Two confounds remain after the arm split. First is the Hawthorne effect: after seeing their dashboard in S1, a with-feedback participant figures out what the system rewards (frequent commits, test runs, IDE actions) and adjusts their behaviour to satisfy it. This looks like improvement, but it is behaviour change to fit the metric, not skill improvement. We can't tell the two apart from the data. Second is learning the task format: all six sessions follow the same structure (a what-not-how prompt, a small change to make, look at the dashboard), so with-feedback participants get better at the format itself, not at refactoring. The baseline arm's downward trend across the same six sessions makes this second explanation harder to sustain in isolation - if format-familiarity alone drove the with-feedback improvement, the baseline participants would also have improved as the format got easier - but it does not formally rule it out. We do not claim either confound is ruled out by the arm split.

\paragraph{The agent extension (\S\ref{sec:results-userstudy-agent}).}
The agent extension is a feasibility experiment for the tool on a structurally different actor, not an agent-monitoring evaluation. Three of the four detector kinds structurally under-fire on agent traces (\S\ref{sec:results-userstudy-agent-limits}): IDE-replay fires on every structural refactoring an agent performs because the agent has no IDE refactor surface to switch to, the commit-gap threshold is rarely reached in a six-session arc, and the edit-burst debouncer collapses each agent turn into one burst. The null arm-level $\Delta J$ result in \S\ref{sec:results-userstudy-agent} reflects detector scope as much as agent behaviour, and would need to be revisited under a detector set reworked for agent edit patterns before any quantitative agent claim could be made. The per-stack engagement split between Claude-family and non-Claude stacks is reported as a qualitative observation from the transcripts, not as a finding about the tool's causal effect. With $n = 6$ vs $n = 2$ across the arm comparison and one stack per axis position in the cross-stack comparison, no inferential weight is given to the per-stack figures.

\paragraph{The rater study (\S\ref{sec:results-kappa}).}
Three labellers, one of whom is us. The chapter reports pairwise $\kappa$ so us-vs-rater values are visible alongside the rater-vs-rater pair, which makes our self-bias detectable rather than hidden. The two participant raters are the same two participants who later performed the user-study sessions, so familiarity with the broader project is not zero; the labels were assigned before any participant performed a user-study session, however, so the labelling cannot have been biased by what the participants later did. Two of the four kinds (\emph{Rework} and \emph{Hygiene}) show $\kappa < 1.0$ with specific disagreement structure named in \S\ref{sec:results-kappa}; reconciliation was deliberately not applied to inflate $\kappa$, since the disagreements are themselves a finding about where the kind boundaries are fuzziest.

\section{Implementation threats}
\label{sec:threats-implementation}

The wrap-and-patch layer of \S\ref{sec:arch-wrap-patch} closes residual gaps between IntelliJ and Eclipse JDT for accepted reorder windows but is not exhaustive: windows where the gap cannot be patched are marked \texttt{ast\_diverged} and excluded from reordering.

The reorder enumerator carries a hard size budget: any window with more than $7$ refactoring steps is skipped entirely, and within an admitted window enumeration stops after $7! = 5040$ permutations (\texttt{TopologicalEnumerator.kt}). The budget exists because the cost of enumerating all topological orderings grows factorially with window size, and an $8$-step window would already require $\geq 40{,}320$ enumerations before the AST-equivalence audit ran on each. In practice the recorded sessions rarely produced reorder windows large enough to exceed this budget, but a session with longer contiguous refactor sequences would have its widest windows silently dropped from the \emph{Ordering} enumeration. A future implementation could replace the exhaustive enumeration with a beam search or heuristic pruning to lift the size cap.

\section{Out of scope}
\label{sec:threats-out-of-scope}

The following are explicitly outside the scope of our work. We do not claim to address them.

\begin{itemize}
    \item \textbf{Agent-side detector rework.} The agent extension of \S\ref{sec:results-userstudy-agent} runs the existing human-trained detectors on eight agent stacks. A quantitative agent evaluation would first require reworking the detectors against an agent-edit model (tool-call grouping in place of edit-burst debouncing, per-turn commit accounting in place of the six-checkpoint window, and an agent-tool-availability check in place of the IDE refactor surface). That work is named in Chapter~\ref{ch:conclusion}.
    \item \textbf{Longitudinal validation against external code-quality outcomes.} No outcome dataset linking refactoring-trajectory signals to downstream maintenance cost was available, and constructing one was treated as outside our scope.
    \item \textbf{Cross-IDE and cross-language validation.} The plugin is IntelliJ-specific and the refactoring engine is JDT-specific (Java). Ports to other IDEs or languages are named as future work and not attempted.
    \item \textbf{Production-weight refitting against an outcome dataset.} See the outcome-dataset point above.
    \item \textbf{Real-time, live-feedback dashboard mode.} The dashboard is a session-replay view of completed analysis reports. Live feedback during refactoring is flagged in \S\ref{sec:results-userstudy} as a natural next step but is not implemented.
    \item \textbf{Rater-rank validation.} As named in \S\ref{sec:threats-internal}, the $\kappa$ study validates kind-presence agreement, not rank-order agreement. A Kendall $\tau$-b study on per-session rankings is the natural follow-up and is not done here.
\end{itemize}
