\documentclass[a4paper, twoside, 11pt]{report}

%% Language and font encodings
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage[T1]{fontenc}

%% Sets page size and margins
\usepackage[a4paper,top=3cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}

%% Useful packages
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{float}
\usepackage{array}
\usepackage{ragged2e}
\usepackage[font=small,labelfont=it,textfont=it]{caption}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, positioning, shapes.geometric, decorations.pathreplacing, calc}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{listings}
\lstdefinestyle{transcript}{
  basicstyle=\ttfamily\scriptsize,
  breaklines=true,
  breakatwhitespace=false,
  columns=fullflexible,
  keepspaces=true,
  showstringspaces=false,
  upquote=true,
  frame=single,
  framesep=4pt,
  xleftmargin=4pt,
  xrightmargin=4pt,
  aboveskip=6pt,
  belowskip=6pt,
}

\newcolumntype{L}[1]{>{\RaggedRight\arraybackslash}p{#1}}

\title{Beyond Before-and-After: Process-Quality Evaluation of Refactoring Sessions TODO: add stuff about the paid tool + user qualitive interviews in appendix}
\author{Ethan Hosier}
% Update supervisor and other title stuff in title/title.tex

\begin{document}
\input{title/title.tex}

\begin{abstract}
This report presents a tool for evaluating refactoring as a process, not just as a change between two endpoints. It defines a process-quality score for refactoring trajectories and uses that score to find points where a developer's path diverges from a higher-scoring alternative. Each divergence is classified by kind, and the system builds a concrete counterfactual reference trajectory for comparison. Developer sessions are captured with an IntelliJ plugin, and the resulting scores and divergences are shown in a dashboard.

The system is evaluated on 45 hand-recorded refactoring sessions on a Java codebase, a 30-session randomised-arm user study on a separate codebase (five participants split into a with-feedback arm of three and a no-feedback baseline arm of two), and an 8-stack, 48-session agent extension on the same fixture as a feasibility experiment for the tool's human-edit-pattern detectors. On the hand-recorded sessions, the tool achieves per-kind precision of 1.00 against inter-rater-agreed labels, with Cohen's $\kappa \geq 0.72$ across the three rater pairs for all four divergence kinds. The score formula's top-$1$ ranking remains stable under single-knob weight perturbations in $96.5\%$ of cases on the user-study rankable subset ($25$ of $30$ sessions), and in $84.6\%$ of cases under simultaneous multi-knob Monte Carlo perturbation. The user-study arm comparison shows a randomised but underpowered direction-of-effect: across the six-session arc the with-feedback arm gains $+22.3$ on the gain-stripped process score (with both the cleanliness-gain and lag weights set to zero) and drives total divergence points from $12$ to $2$, while the no-feedback baseline arm loses $2.0$ and rises from $10$ to $15$ divergence points. The agent extension finds no equivalent feedback effect, consistent with structural biases of the human-trained detectors against agent edit patterns.
\end{abstract}

\tableofcontents

\input{introduction/introduction.tex}
\input{background/background.tex}
\input{methodology/methodology.tex}
\input{results/results.tex}
\input{threats/threats.tex}
\input{conclusion/conclusion.tex}
\input{declarations/declarations.tex}
\input{appendix/appendix.tex}

\bibliographystyle{unsrt}
\bibliography{bibs/sample}

\end{document}